Efficient ancestry and mutation simulation with msprime 1.0

### 2021-11-17 response-to-reviewers.tex

\begin{point} Figure comments
Figure 2
I think it could be made more clear that the genotype matrix is not part of the set of tables.
\end{point}

\begin{reply}
We agree, and have revised Figure 2 and the caption to clarify the distinction between
stored and derived data.
\end{reply}

\begin{point}
Figure 4
It is not clear if these two plots present all the data used for the quadratic
fits, or if these data continue outside of the plots. Samples are described,
but not their ploidy, are they haploids or diploid? From the plots here, it is
not clear if the quadratic relationship holds for sample sizes << 1000. I
suggest this figure should also include a smaller sample size, (eg n=10)
\end{point}

\begin{reply}
We have moved Figure 4 to the appendix, as it is of more specialised interest
than the other main text figures. We have truncated the x-axis on the
quadratic fits to just the available data, to clarify that we are showing
all of the data. We have clarified that samples are diploid.

The plot is already quite complex and visually busy, and it is difficult to
see how we could add a third line without major changes. The figure is
primarily intended as a useful yardstick for users who wish to run simulations
and would like a rough idea of how long it should take, and therefore adding
a third line for very small sample sizes is likely to make it less useful
for this purpose.
\end{reply}

### 2021-11-17 paper.tex
Respond to Hein et al point

A major change for the \msprime\ 1.0 release is the introduction of a new set of APIs,
designed in part to avoid sources of error (see the Demography section) but
also to provide more appropriate defaults while keeping compatibility with
existing code. In the new APIs, ancestry and mutation simulation are fully
separated (see Fig.~\ref{fig-mutated-trees}),
with the \texttt{sim\_ancestry} and \texttt{sim\_mutations}
functions replacing the legacy \texttt{simulate} function. Among other changes,
the new APIs default to discrete genome coordinates and finite sites mutations,
making the default settings more realistic and resolving a major source of confusion and error.
The previous APIs are fully
supported and tested, and will be maintained for the foreseeable future.
The \texttt{msp} program has been extended to include new commands
for simulating ancestry and mutations separately. A particularly useful
feature is the ability to specify demographic models in
Demes format~\citep{gower2021demes} from the command line,
making simulation of complex demographies straightforward.
We also provide an \ms\ compatible
command line interface to support existing workflows.

@@

As discussed in the \nameref{sec-recombination} section, the time complexity of
Hudson's algorithm is quadratic in the population scaled recombination rate
$\rho = 4 N_e L$, where $L$ is the length of the genome in units of
recombination distance. For large $\rho$, recombination events in trapped
ancestral material will dominate, and so we can use this as a proxy for the
overall number of events in Hudson's algorithm.
\citet[Eq.~5.10]{hein2004gene} gave
\begin{equation}\label{eqn-hsw-bound}
\rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2
\end{equation}
as an upper bound
on the number of recombination events within trapped ancestral material
in Hudson's algorithm for $n$ samples.
% Since the sum is well approximated by $\log{n}$,
% a natural conjecture is that the time complexity of Hudson's
% algorithm is $O(\rho^2 \log^2 n)$.
Fig.~\ref{fig-ancestry-perf} shows the observed run time for
simulations with a variety of population size, chromosome
length and sample sizes, and demonstrates that Eq.~\eqref{eqn-hsw-bound}
correctly predicts the quadratic dependence on $\rho$,
as previously conjectured~\citep[Fig.~2]{kelleher2016efficient}.
We also see that the dependence on $n$ is quite weak,
since increasing sample size 100-fold
only increases run time by a factor of 2 or so. However, the
$\log^2{n}$ factor implied by Eq.~\eqref{eqn-hsw-bound}
(the sum is a harmonic number and can be approximated by $\log{n}$)
is not well supported by observed run times (or numbers of events)
except possibly at very
large values of $\rho$. It therefore appears that a different dependence
on $n$ is required to accurately predict simulation time for a
given $\rho$ and $n$.

>>>>

A major change for the \msprime\ 1.0 release is the introduction of a new set of APIs,
designed in part to avoid sources of error (see the Demography section) but
also to provide more appropriate defaults while keeping compatibility with
existing code. In the new APIs, ancestry and mutation simulation are fully
separated (see Fig.~\ref{fig-mutated-trees}),
with the \texttt{sim\_ancestry} and \texttt{sim\_mutations}
functions replacing the legacy \texttt{simulate} function. Among other changes,
the new APIs default to discrete genome coordinates and finite sites mutations,
making the default settings more realistic and resolving a major source of confusion and error.
The previous APIs are fully
supported and tested, and will be maintained for the foreseeable future.
The \texttt{msp} program (a command line interface to the library)
has been extended to include new commands
for simulating ancestry and mutations separately. A particularly useful
feature is the ability to specify demographic models in
Demes format~\citep{gower2021demes} from the command line,
making simulation of complex demographies straightforward.
We also provide an \ms\ compatible
command line interface to support existing workflows.

@@

As discussed in the \nameref{sec-recombination} section, the time complexity of
Hudson's algorithm is quadratic in the population scaled recombination rate
$\rho = 4 N_e L$, where $L$ is the length of the genome in units of
recombination distance. For large $\rho$, recombination events in trapped
ancestral material will dominate, and so we can use this as a proxy for the
overall number of events in Hudson's algorithm.
\citet[Eq.~5.10]{hein2004gene} gave
\begin{equation}\label{eqn-hsw-bound}
\rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2
\end{equation}
as an upper bound
on the number of recombination events within trapped ancestral material
in Hudson's algorithm for $n$ samples.
% Since the sum is well approximated by $\log{n}$,
% a natural conjecture is that the time complexity of Hudson's
% algorithm is $O(\rho^2 \log^2 n)$.
Fig.~\ref{fig-ancestry-perf} shows the observed run time for
simulations with a variety of population size, chromosome
length and sample sizes, and demonstrates that Eq.~\eqref{eqn-hsw-bound}
correctly predicts the quadratic dependence on $\rho$
\citep[see also][Fig.~2]{kelleher2016efficient}.
We also see that the dependence on $n$ is quite weak,
since increasing sample size 100-fold
only increases run time by a factor of 2 or so. However, the
$\log^2{n}$ factor implied by Eq.~\eqref{eqn-hsw-bound}
(the sum is a harmonic number and can be approximated by $\log{n}$)
is not well supported by observed run times (or numbers of events)
except possibly at very
large values of $\rho$. It therefore appears that a different dependence
on $n$ is required to accurately predict simulation time for a
given $\rho$ and $n$.
